Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-16 _paper.rmd
- Add chocolates inverse case. - ames explaintion++ - Comb over explanations and variable formating - checkin before changing SHAP_bd plot adding sequences

<!-- SHAP -->
SHaply Additive exPlanations (SHAP) quantifies the feature contributions of one instance, by examining the effect of other features on the predictions. The explanations of SHAP almost all refer to @shapley_value_1953's method to evaluate an individual's contribution to cooperative games by assessing the performance of this player in the presence or absence of other players. @strumbelj_efficient_2010 introduced the use of SHAP for local explanations in ML models. <!--While the models themselves are invariant to the order of the predictors, the additive explanations are not.--> The SHAP values are the mean contributions over some subset of permutations of features. The approach is  related to partial dependence plots [@molnar_interpretable_2020], used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015, which helps to understand the relative importance of each feature in the context of all other candidate features.

@@

(ref:casepenguins) Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is an Chinstrap (orange) penguin that is misclassified as a Gentoo (purple), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of `f_l` from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably a Chinstrap than a Gentoo (b, right). (The .mp4 animation of the radial tour is at [shorturl.at/hFHNS](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4).)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins)"}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

Figure \@ref(fig:casepenguinsblfl) confirms that flipper length (`f_l`) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and we can see that the PI is closer to the Chinstrap group in these two features, mostly because it has an unusually low value of flipper length relative to other Gentoos.

@@

(ref:casechocolates) Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove four features with the lowest contribution for the selected instance and vary the contribution of `fiber`. The misclassification of two instances seems improbable when sugar is near the max contribution. Animated tour can be found at [shorturl.at/dBE28](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4).
```{r casechocolates, echo=F, out.width="100%", fig.cap = "(ref:casechocolates)"}
knitr::include_graphics("./figures/case_chocolates.png")
```

@@

<!-- Figure \@ref(fig:casechocolates) similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. We look at one extreme instance and vary the contribution of `fiber`. We see notice two instances that have clearly well within the milk chocolate despite being observed as dark chocolate on their product label. While conjecture, there are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. In the penguin case we explored how the model misclassified an instance from a type one error perspective. In contrast, in the chocolates cases shows a type two error view; looking for extreme support of the models classification to challenge the original instance.-->

@@

(ref:casefifa) FIFA 2020 data, a random forest model, regresses wages [2020 Euros] from nine aggregated skill measurements. The PI is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). We remove three features with low attribution from both players. The attribution projection starts with the selected instance on the right. We vary the contribution from `defense`, the star offensive player is not distinguished in the horizontal direction. At this point, defensive players have been rotated to the highest horizonatal value. The animate radial tour can be found at [shorturl.at/lCK26](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4)
```{r casefifa, echo=F, out.width = "100%", fig.cap = "(ref:casefifa)"}
knitr::include_graphics("./figures/case_fifa.png")
```

With figure \@ref(fig:casefifa), we will test the premise of the local explanation. While `offense` and `reaction` skills are both crucial to explaining a star offensive player. As the contribution of defensive skills increases, Messi's is no longer separated from the group, and other defensive players are better predicted in this attribution case. In terms of what-if analysis, were Messi's tree SHAP attributions at these levels, his predicted wages would be halved.

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model has regressed this price with the features `LotArea` (`LtA`), `OverallQual` (`Qlt`), `YearBuild` (`YrB`), `LivingArea` (`LvA`), `Bathrms` (`Bth`), `Bedrms` (`Bdr`), `TotRms` (`Rms`), `GarageYrBlt` (`GYB`), and `GarageArea` (`GrA`). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

(ref:caseames) Ames housing 2018 regressing sales price [USD]. The PI sale price was under predicted, and had sizable attribution to lot area (`LtA`). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of `LtA`,  which is important to explaining the PI and near invariant to the sales price of the CI. The corresponding animation is at [shorturl.at/oqyDU](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

Figure \@ref(fig:caseames) selects the house sale 74, a sizeable under prediction that has a large contribution to lot area. The CI has a similar predicted price though the prediction was accurate and gives almost no attribution to lot size. As the contribution of lot area decreases, the predictive power decreases for the PI, while the CI remains stationary.

>>>>

<!-- SHAP -->
SHaply Additive exPlanations (SHAP) quantifies the feature contributions of one instance, by examining the effect of other features on the predictions. The explanations of SHAP almost all refer to @shapley_value_1953's method to evaluate an individual's contribution to cooperative games by assessing the performance of this player in the presence or absence of other players. @strumbelj_efficient_2010 introduced the use of SHAP for local explanations in ML models. The attribution of feature importance depends on the seqence of the features already included. The SHAP values are the mean contributions over different sequences of the of features. The approach is related to partial dependence plots [@molnar_interpretable_2020], used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in @wickham_visualizing_2015, which helps to understand the relative importance of each feature in the context of all other candidate features.

@@

(ref:casepenguins) Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is an Chinstrap (orange) penguin that is misclassified as a Gentoo (purple), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of `f_l` from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at [shorturl.at/hFHNS](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4).)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins)"}
knitr::include_graphics("./figures/case_penguins.png")
```

@@

Figure \@ref(fig:casepenguinsblfl) confirms that flipper length (`f_l`) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and we can see that the PI is closer to the Chinstrap group in these two features, mostly because it has an unusually low value of flipper length relative to other Gentoos. From this view is makes sense that is a hard instance to account for as decsion trees can only partition only vertical and horizontal lines.

@@

(ref:casechocolates) Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove four features with the lowest contribution for the selected instance and vary the contribution of Fiber. The misclassification seems improbable when sugar is near the max contribution. Animated tour can be found at [shorturl.at/dBE28](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4).
```{r casechocolates, echo=F, out.width="100%", fig.cap = "(ref:casechocolates)"}
knitr::include_graphics("./figures/case_chocolates.png")
```

@@

(ref:casechocolatesinverse) Chocolates data type classification (milk or dark). Looking at the inverse misclassification, we select a milk chocolate while the model predicts it to be dark chocolate. Sodium and Fiber (`Na` and `Fbr`) have the largest differences in attributed feature importance. We remove features with the lowest contributions and vary the contribution of Sodium. This misclassification is not supported when sodium has contribution close to the attribution alligned with the other milk chocolates. Animated tour can be found at [shorturl.at/dBE28](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4).
```{r casechocolatesinverse, echo=F, out.width="100%", fig.cap = "(ref:casechocolatesinverse)"}
knitr::include_graphics("./figures/case_chocolates_inverse.png")
```

It would also be interesting to explore the inverse case; which features lead to a mild chocolate being misclassified as dark. Chocolate 84 is just this case and we compare with a correctly predicted milk chocolate (instance 71). This exploration is shown in Figure \@ref(fig:casechocolatesinverse). From the parallel coordinate lines, we identidy discrepancies in Sodium and Fiber. We opt to vary Sodium, and find the attribtuion of the other milk chocolates not to support the prediction of a dark chocolate.

@@

(ref:casefifa) FIFA 2020 data, a random forest model, regresses wages [2020 Euros] from nine aggregated skill measurements. The PI is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). We remove three features with low attribution from both players. The attribution projection starts with the selected instance on the right. We vary the contribution from defense (`def`), the star offensive player is not distinguished in the horizontal direction. At this point, defensive players have been rotated to the highest horizonatal value. The animate radial tour can be found at [shorturl.at/lCK26](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4).
```{r casefifa, echo=F, out.width = "100%", fig.cap = "(ref:casefifa)"}
knitr::include_graphics("./figures/case_fifa.png")
```

With figure \@ref(fig:casefifa), we will test the premise of the local explanation. While offensive and reaction skills (`off` and `rct`) are both crucial to explaining a star offensive player. As the contribution of defensive skills increases, Messi's is no longer separated from the group, and other defensive players are better predicted in this attribution case. In terms of what-if analysis, were Messi's tree SHAP attributions at these levels, his predicted wages would be halved.

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model has regressed this price with the features Lot Area (`LtA`), Overall Qualality (`Qlt`), Year the house was Build (`YrB`), Living Area (`LvA`), number of Bathrooms (`Bth`), number of Bedrooms (`Bdr`), total number of Rooms (`Rms`), Year the Garage was Build (`GYB`), and Garage Area (`GrA`). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

(ref:caseames) Ames housing 2018 regressing sales price [USD]. The PI sale price was under predicted, and had sizable attribution to lot area (`LtA`). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of `LtA`, which is important to explaining the PI and near invariant to the sales price of the CI. The corresponding animation is at [shorturl.at/oqyDU](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4).
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

Figure \@ref(fig:caseames) selects the house sale 74, a sizable under prediction that has a large contribution to lot area. The CI has a similar predicted price though the prediction was accurate and gives almost no attribution to lot size. The attribution projection is places instances with high living areas to the right. We control the contribution of this feature. As the contribution of lot area decreases, the predictive power decreases for the PI, while the CI remains stationary. This large of an importance is living area is relatively uncommon. Boosting tree models may be more resilient to such a under prediction as up-weighting this residual would force its inclusion in the final model.

### 2022-01-16 _paper.tex
- Add chocolates inverse case. - ames explaintion++ - Comb over explanations and variable formating - checkin before changing SHAP_bd plot adding sequences

SHaply Additive exPlanations (SHAP) quantifies the feature contributions of one instance, by examining the effect of other features on the predictions. The explanations of SHAP almost all refer to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by assessing the performance of this player in the presence or absence of other players. Strumbelj and Kononenko (2010) introduced the use of SHAP for local explanations in ML models. The SHAP values are the mean contributions over some subset of permutations of features. The approach is related to partial dependence plots (Molnar 2020), used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each feature in the context of all other candidate features.

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is an Chinstrap (orange) penguin that is misclassified as a Gentoo (purple), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of \texttt{f\_l} from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably a Chinstrap than a Gentoo (b, right). (The .mp4 animation of the radial tour is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{shorturl.at/hFHNS}.)}\label{fig:casepenguins}
\end{figure}

@@

Figure \ref{fig:casepenguinsblfl} confirms that flipper length (\texttt{f\_l}) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and we can see that the PI is closer to the Chinstrap group in these two features, mostly because it has an unusually low value of flipper length relative to other Gentoos.

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove four features with the lowest contribution for the selected instance and vary the contribution of \texttt{fiber}. The misclassification of two instances seems improbable when sugar is near the max contribution. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{shorturl.at/dBE28}.}\label{fig:casechocolates}
\end{figure}

@@

@@

\caption{FIFA 2020 data, a random forest model, regresses wages {[}2020 Euros{]} from nine aggregated skill measurements. The PI is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). We remove three features with low attribution from both players. The attribution projection starts with the selected instance on the right. We vary the contribution from \texttt{defense}, the star offensive player is not distinguished in the horizontal direction. At this point, defensive players have been rotated to the highest horizonatal value. The animate radial tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4}{shorturl.at/lCK26}}\label{fig:casefifa}
\end{figure}

With figure \ref{fig:casefifa}, we will test the premise of the local explanation. While \texttt{offense} and \texttt{reaction} skills are both crucial to explaining a star offensive player. As the contribution of defensive skills increases, Messi's is no longer separated from the group, and other defensive players are better predicted in this attribution case. In terms of what-if analysis, were Messi's tree SHAP attributions at these levels, his predicted wages would be halved.

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model has regressed this price with the features \texttt{LotArea} (\texttt{LtA}), \texttt{OverallQual} (\texttt{Qlt}), \texttt{YearBuild} (\texttt{YrB}), \texttt{LivingArea} (\texttt{LvA}), \texttt{Bathrms} (\texttt{Bth}), \texttt{Bedrms} (\texttt{Bdr}), \texttt{TotRms} (\texttt{Rms}), \texttt{GarageYrBlt} (\texttt{GYB}), and \texttt{GarageArea} (\texttt{GrA}). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

@@

\caption{Ames housing 2018 regressing sales price {[}USD{]}. The PI sale price was under predicted, and had sizable attribution to lot area (\texttt{LtA}). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of \texttt{LtA}, which is important to explaining the PI and near invariant to the sales price of the CI. The corresponding animation is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{shorturl.at/oqyDU}}\label{fig:caseames}
\end{figure}

Figure \ref{fig:caseames} selects the house sale 74, a sizeable under prediction that has a large contribution to lot area. The CI has a similar predicted price though the prediction was accurate and gives almost no attribution to lot size. As the contribution of lot area decreases, the predictive power decreases for the PI, while the CI remains stationary.

>>>>

SHaply Additive exPlanations (SHAP) quantifies the feature contributions of one instance, by examining the effect of other features on the predictions. The explanations of SHAP almost all refer to Shapley (1953)'s method to evaluate an individual's contribution to cooperative games by assessing the performance of this player in the presence or absence of other players. Strumbelj and Kononenko (2010) introduced the use of SHAP for local explanations in ML models. The attribution of feature importance depends on the seqence of the features already included. The SHAP values are the mean contributions over different sequences of the of features. The approach is related to partial dependence plots (Molnar 2020), used to explain the effect of a feature by predicting the response for a range of values on this feature, after fixing the value of all other features to their mean. Partial dependence plots are a global approximation of the feature importance, while SHAP is specific to one instance. It could also be considered to be similar to examining the coefficients from all subsets regression, as described in Wickham, Cook, and Hofmann (2015), which helps to understand the relative importance of each feature in the context of all other candidate features.

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is an Chinstrap (orange) penguin that is misclassified as a Gentoo (purple), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of \texttt{f\_l} from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably a Chinstrap than a Gentoo (b, right). (The animation of the radial tour is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{shorturl.at/hFHNS}.)}\label{fig:casepenguins}
\end{figure}

@@

Figure \ref{fig:casepenguinsblfl} confirms that flipper length (\texttt{f\_l}) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted, and we can see that the PI is closer to the Chinstrap group in these two features, mostly because it has an unusually low value of flipper length relative to other Gentoos. From this view is makes sense that is a hard instance to account for as decsion trees can only partition only vertical and horizontal lines.

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove four features with the lowest contribution for the selected instance and vary the contribution of Fiber. The misclassification seems improbable when sugar is near the max contribution. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{shorturl.at/dBE28}.}\label{fig:casechocolates}
\end{figure}

@@

{\centering \includegraphics[width=1\linewidth]{./figures/case_chocolates_inverse}

\caption{Chocolates data type classification (milk or dark). Looking at the inverse misclassification, we select a milk chocolate while the model predicts it to be dark chocolate. Sodium and Fiber (\texttt{Na} and \texttt{Fbr}) have the largest differences in attributed feature importance. We remove features with the lowest contributions and vary the contribution of Sodium. This misclassification is not supported when sodium has contribution close to the attribution alligned with the other milk chocolates. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{shorturl.at/dBE28}.}\label{fig:casechocolatesinverse}
\end{figure}

It would also be interesting to explore the inverse case; which features lead to a mild chocolate being misclassified as dark. Chocolate 84 is just this case and we compare with a correctly predicted milk chocolate (instance 71). This exploration is shown in Figure \ref{fig:casechocolatesinverse}. From the parallel coordinate lines, we identidy discrepancies in Sodium and Fiber. We opt to vary Sodium, and find the attribtuion of the other milk chocolates not to support the prediction of a dark chocolate.

@@

\caption{FIFA 2020 data, a random forest model, regresses wages {[}2020 Euros{]} from nine aggregated skill measurements. The PI is a star offensive player (L. Messi) compared with a top defensive player (V. van Dijk). We remove three features with low attribution from both players. The attribution projection starts with the selected instance on the right. We vary the contribution from defense (\texttt{def}), the star offensive player is not distinguished in the horizontal direction. At this point, defensive players have been rotated to the highest horizonatal value. The animate radial tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_fifa.mp4}{shorturl.at/lCK26}.}\label{fig:casefifa}
\end{figure}

With figure \ref{fig:casefifa}, we will test the premise of the local explanation. While offensive and reaction skills (\texttt{off} and \texttt{rct}) are both crucial to explaining a star offensive player. As the contribution of defensive skills increases, Messi's is no longer separated from the group, and other defensive players are better predicted in this attribution case. In terms of what-if analysis, were Messi's tree SHAP attributions at these levels, his predicted wages would be halved.

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest model has regressed this price with the features Lot Area (\texttt{LtA}), Overall Qualality (\texttt{Qlt}), Year the house was Build (\texttt{YrB}), Living Area (\texttt{LvA}), number of Bathrooms (\texttt{Bth}), number of Bedrooms (\texttt{Bdr}), total number of Rooms (\texttt{Rms}), Year the Garage was Build (\texttt{GYB}), and Garage Area (\texttt{GrA}). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

@@

\caption{Ames housing 2018 regressing sales price {[}USD{]}. The PI sale price was under predicted, and had sizable attribution to lot area (\texttt{LtA}). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of \texttt{LtA}, which is important to explaining the PI and near invariant to the sales price of the CI. The corresponding animation is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{shorturl.at/oqyDU}.}\label{fig:caseames}
\end{figure}

Figure \ref{fig:caseames} selects the house sale 74, a sizable under prediction that has a large contribution to lot area. The CI has a similar predicted price though the prediction was accurate and gives almost no attribution to lot size. The attribution projection is places instances with high living areas to the right. We control the contribution of this feature. As the contribution of lot area decreases, the predictive power decreases for the PI, while the CI remains stationary. This large of an importance is living area is relatively uncommon. Boosting tree models may be more resilient to such a under prediction as up-weighting this residual would force its inclusion in the final model.
